\Large{All in the family:\\Anna and Lisa Hahner's finishing times in\\the 2016 women's Olympic marathon}

### 2017-01-29 paper/paper.tex
other edits

\begin{quote}
 \emph{``I invested all I had and 300 meters before the finish line, I was next to Lisa. It was a magical moment that we could finish this marathon together. We did not think about what we were doing.''€ - Anna Hahner} 
\end{quote}

@@

@@

Thomas Kurschilgen's accusations of an intentional finish stems from
two observations. First, the Hahner twins finished slower than he
expected, as we already noted recording times more than 18 minutes
slower than respective personal bests. Second, the twins finished
together at the exactly same time. Kurschilgen clearly believed that
neither of these events would have occurred had both Hahner run the
marathon independently, absent coordination. By his logic, the German
twins should have run faster and not have finished simultaneously.

On the other hand, only a handful of runners completed the Rio
marathon with times that were faster than their recorded best times.
Thus, 18 minutes behind a personal record may not be the outlier that
Kurschilgen claimed it to be. Moreover, there is good reason to
suspect that an unintended simultaneous finish was more likely for the
Hahner twin than it would be for any other set of runners. After all,
these two women are twins with presumably similar abilities. Not only
do they train together, but their pre-Rio best times are less than two
minutes apart. While Anna may be slightly faster than Lisa, measured
by personal bests, we might expect the difference in their Rio result
to be just as close as the differences in their recorded bests. And
given random variation in finishing times, a simultaneous finish might
not be out of the ordinary.

@@

To test the assertion that the Hahner twins paced themselves to finish
at the same time and with back-to-back placements, we simply need to
compute the probability that such a result would have occurred
unintentionally. Hence we need to know the probability distribution of
Anna and Lisa's finishing times if their runs had been independent of
each other. The challenge, of course, is that this distribution is
unknown.  ^^ Therefore it needs to be estimated.

@@

%% Ok.  Not sure what all the notation means.

@@

Suppose that Hahner the twins acted independently $I$ or they acted in
coordination $C$; these are the two possible states of the race
$\psi = \{I, C\}$. Given that we observed Anna and Lisa's finishing
times, $Y_A = y_A$ and $Y_L = y_L$, we want to know the probability
that they ran independently, as they say they did. Hence, we are
looking to determine,  %% y_a versus Y_a.  Former realized?

@@

\begin{figure}[!ht]
  \caption{Relationship between Personal Best and Result  -- better label?}
  \label{fig:scatter}
  \begin{subfigure}{.5\textwidth}
    \includegraphics[width=\textwidth,
    keepaspectratio]{scatter_plot.pdf}
    \caption{Rio finishing times and personal best times}
    \label{fig:A}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
    % \label{fig:Studentized Residuals}
    \includegraphics[width=.975\textwidth, keepaspectratio]{studentized_residuals.pdf}
    \caption{A}
    \label{fig:B}
  \end{subfigure}
\end{figure}

The scatter plot in Figure \ref{fig:scatter} displays the relationship between a runner's personal best and their final result.   Of the 156 runners that competed in the marathon event, 133 finished the race.  These individuals are plotted as grey points.  The 23 that did not finish are plotted below along the x-axis as grey tick-marks.   The dashed black line with grey confidence intervals represent OLS estimates of the linear relationship between the two variables.   The Hahner twins are marked by orange points.

One can see that there is a clear linear relationship between a runner's performance in the olympics and her final result.  Moreover, there is a significant amount of noise in the relationship.  This suggests that despite one's innate abilities, luck plays a major role in olympic finishes.    Still, the twins' results are well above the fitted line, even accounting for typical error.  They are much slower than the underlying linear relationship suggests.  In fact, if we order each point by the size of the studentized residual (see the second plot), we can see that the Hahner twins are in the tail-end of the distribution.  Anna Hahner's residual deviation is greater than 91.7\% of the runners who completed the race while Lisa Hahner's deviation is greater than 86.5\%.  Hence, the twins appear to have finished at a much slower pace than expectation, which is what we might expect if they were coordinating their runs.\footnote{It is important to note that the North Korean twins finished much faster than expected.  XXXX explain PRK}

@@

\begin{enumerate}
\item Removing the twins from the sample, estimate a linear model that
predicts a runner?s final time ($Y_i$) based on her personal best time ($X_i$).
\item Extract the coefficients and the variance-covariance matrix from this
model.
\item For each runner, draw $\beta_0$ and $\beta_1$ from a multivariate normal
distribution with mean equal to the coefficients and variance equal to the
variance-covariance matrix.
\item For each runner, draw an error term ($e$) from a normal distribution with
mean equal to zero and a standard deviation equal to the standard
deviation of the model?s residuals ($\sigma$).
\item For each runner, predict the final result by combining the randomly
generated beta coefficients and error terms, $\hat{y} = \beta_0 + \beta_1 + e$.
\item Eliminate each runner from the race with a probability equal to the
percent of the total runners who did not finish in the actual race.  This
account for the likelihood of a DNF.
\item Calculate the difference in time between Anna Hahner and Lisa Hahner.
\item Calculate the difference in ranking between Anna Hahner and Lisa Hahner.
\item Repeat steps 3 through 8 ten thousand times.
\item Plot a histogram of the results in Figure \ref{fig:simdiff}
\end{enumerate}

\begin{figure}[!ht]
 \caption{The distribution of simulated results}
 \label{fig:simdiff}
 \begin{subfigure}{.5\textwidth}
 \includegraphics[width=\textwidth, keepaspectratio]{simulated_time.pdf}
 \end{subfigure}
 \begin{subfigure}{.5\textwidth}
 \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank.pdf}
 \end{subfigure}
\end{figure}

>>>>

\begin{quote}
  \emph{``I invested all I had and 300 meters before the finish line,
    I was next to Lisa. It was a magical moment that we could finish
    this marathon together. We did not think about what we were
    doing.'' -- Anna Hahner}
\end{quote}

@@

As a brief overview of the race, Figure \ref{fig:secondsbehind}
describes each runner's status at the various split times on the
marathon course.  Each dot in the figure depicts a recorded split and
the number of seconds each runner was behind the race leader at the
time.  There are more dots at earlier splits due to the accumulation
of DNFs.  Regarding twins and triplets, the Estonian Luik triplets are
denoted with blue dots; there are three blue dots through the 20
kilometer split, but the third triplet did not record a half-marathon
split.  The orange dots in the figure represent the Hahner twins; Lisa
Hahner was ahead of her sister through 15 kilometers at which point
Anna surged ahead.  There are two red dots representing the North
Korean Kim twins although this is not visually apparent; the Kim twins
had identical split times during the marathon.

\begin{figure}[!ht]
  \centering
  \caption{Runner status by split}
  \label{fig:secondsbehind}
  \includegraphics[scale = 1]{seconds-behind.pdf}
  \begin{flushleft}
    \emph{Note: each dot represents one runner at a split.  DNFs are
      not pictured, and splits are not to scale.}
  \end{flushleft}
\end{figure}

Figure \ref{fig:secondsbehind} does not directly shed any light on our
motivating research question, but it does show that the Hahner twins
finished amidst many other runners.  We will come back to this point
later.

@@

Thomas Kurschilgen's accusations of an intentional finish stems from
two observations. First, the Hahner twins finished slower than he
expected given respective personal bests. Second, the twins finished
together at the exactly same time. Kurschilgen clearly believed that
neither of these events would have occurred had both Hahner run the
marathon independently, absent coordination. By his logic, the German
twins should have run faster and not have finished simultaneously.

On the other hand, only a handful of runners completed the Rio
marathon with times that were faster than their recorded best times.
Thus, 18 minutes behind a personal record may not be the outlier that
Kurschilgen claimed it to be. Moreover, there is good reason to
suspect that an unintended simultaneous finish was more likely for the
Hahner twin than it would be for any other set of runners. After all,
these two women are twins with presumably similar abilities. Not only
do they train together, but their pre-Rio best times are less than two
minutes apart. While Anna may be slightly faster than Lisa measured by
personal bests, we might expect the difference in their Rio result to
be just as close as the differences in their recorded bests. And given
random variation in finishing times, a simultaneous finish might not
be out of the ordinary.

@@

To test the assertion that the Hahner twins paced themselves to finish
at the same time and with back-to-back placements, we need to compute
the probability that such a result would have occurred
unintentionally. Hence we need to know the probability distribution of
Anna and Lisa's finishing times if their runs had been independent of
each other. The challenge, of course, is that this distribution is
unknown.  % Therefore it needs to be estimated.

@@

%% Ok!  Not sure what all the notation means.

@@

The Hahner twins either acted independently $I$ or coordinated their
times $C$; these are the two possible states of the race
$\psi = \{I, C\}$. Given that we observed Anna and Lisa's finishing
times, $Y_A = y_A$ and $Y_L = y_L$, we want to know the probability
that they ran independently, as they say they did. Hence, we are
looking to determine, %% y_a versus Y_a.  Former realized?

@@

\begin{figure}[!ht]
  \caption{Relationship between Personal Best and Result  -- caption?}
  \label{fig:scatter}
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth,
    keepaspectratio]{scatter_plot.pdf}
    \caption{Rio finishing times and personal best times}
    \label{fig:45degreeplot}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
    \centering
    % \label{fig:Studentized Residuals}
    \includegraphics[width=.975\textwidth, keepaspectratio]{studentized_residuals.pdf}
    \caption{Studentized residuals}
    \label{fig:studentizedresiduals}
  \end{subfigure}
\end{figure}

Figure \ref{fig:45degreeplot} displays the relationship between a
runner's personal best (horizontal axis) and her final result
(vertical axis).  The 133 finishers are plotted as grey points, and
the 23 DNFs are depicted below along the horizontal axis with
tick-marks.  The dashed black line with grey confidence intervals
represent ordinarily least squares estimates of the linear
relationship between racer's finishing and personal best times.

One can see in Figure \ref{fig:scatter} that there is a clear
relationship between a runner's personal best time and her performance
in the Olympics.  While there is a significant amount of noise in the
relationship, the Hahner twins' dots are well above the regression's
fitted line notwithstanding this point.  They finished the Rio
marathon much slower than they should have, per their pre-Olympic
personal best times.  In fact, if we sort studentized residual, we can
see in Figure \ref{fig:studentizedresiduals} that the Hahner twins are
in the tail-end of the residual distribution.  Anna Hahner's residual
deviation is greater than 91.7\% of the runners who completed the
marathon while Lisa Hahner's deviation is greater than 86.5\%.  Hence,
the twins appear to have finished at a much slower pace than expected,
which is what we might expect if they were coordinating their runs.
The North Korean Kim twins are also in the tail of the residual
distribution---but the left tail.  These twins finished much faster
than expected.

@@

\begin{enumerate}
\item Removing the two sets of twins and triplets from the sample,
  estimate a linear model that predicts a runner's final time ($Y_i$)
  based on her pre-Olympic personal best time ($X_i$).
\item Extract the coefficients and the variance-covariance matrix from
  this model.
\item For each runner, draw $\beta_0$ and $\beta_1$ from a multivariate normal
distribution with mean equal to the coefficients and variance equal to the
variance-covariance matrix.
\item For each runner, draw an error term ($e$) from a normal distribution with
mean equal to zero and a standard deviation equal to the standard
deviation of the model's residuals ($\sigma$).
\item For each runner, predict the final result by combining the randomly
generated beta coefficients and error terms, $\hat{y} = \beta_0 + \beta_1 + e$.
\item Eliminate each runner from the race with a probability equal to the
percent of the total runners who did not finish in the actual race.  This
account for the likelihood of a DNF.
\item Calculate the difference in time between Anna Hahner and Lisa Hahner.
\item Calculate the difference in ranking between Anna Hahner and Lisa Hahner.
\item Repeat steps 3 through 8 ten thousand times.
\item Plot a histogram of the results in Figure \ref{fig:simdiff}
\end{enumerate}

\begin{figure}[!ht]
  \centering
  \caption{Distribution of simulated results for the Hahner twins}
  \label{fig:simdiff}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth,
    keepaspectratio]{simulated_time.pdf}
    \caption{Simulated differences in finishing times}
    \label{fig:simulatedfinishtimes}
  \end{subfigure}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank.pdf}
    \caption{Simulated differences in places}
    \label{fig:simulatedranks}
  \end{subfigure}
\end{figure}
